---
title: "Load data, estimate and compare models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimate a model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Importing a `.mom` file as a `gnssts` object



Let us first load the `gmwmx` package.
```{r, warning=F, message=F}
library(gmwmx)
```

```{r, echo = F, eval=F}
dobs = gmwmx::PBO_get_station("DOBS", column = "dN")
write.gnssts(dobs, filename = "data_dobs.mom")
```

Consider that you want to estimate a model on data saved in a `.mom` file located at a specific `file_path` on your computer. 
```{r, echo=F}
file_path = system.file("extdata", "data_dobs.mom", package = "gmwmx", mustWork = T)
data_dobs = read.gnssts(filename = file_path)
```


For example, the corresponding `.mom` file could have a similar looking:

```
# sampling period 1.000000
# offset 55285.000000
# offset 58287.770833
52759.5 -0.01165
52760.5 -0.01102
52761.5 -0.01147
...
```

You can import the `.mom` file as a with the function `read.gnssts()` as such:

```{r}
data_dobs = read.gnssts(filename = file_path)
```

Objects created or imported with `create.gnss()` or `read.gnssts()` are of class `gnssts`.
```{r}
class(data_dobs)
```


By inspecting the structure of a `gnssts` object, we observe that `gnssts` objects specify the time vector, the observation vector, the sampling period and the times at which there are location shifts (jumps).
```{r}
str(data_dobs)
```


We can represent the signal as such:

```{r, fig.height=5, fig.align='center', fig.width=6}
plot(data_dobs$t, data_dobs$y, type="l")
```


# Estimate a model 

## Supported models
The `gmwmx` package allows to estimate linear model with correlated residuals that are described by a functional model and a stochastic noise model.

### Functional model

More precisely, for the functional model, we consider a linear model which can be expressed as:

\begin{equation}
    \mathbf{Y} = \mathbf{A} {{\bf x}}_0 + \boldsymbol{\varepsilon},

\end{equation}

where $\mathbf{Y} \in {\rm I\!R}^n$ denotes the response variable of interest (i.e., vector of GNSS observations), $\mathbf{A} \in {\rm I\!R}^{n \times p}$ a fixed design matrix, ${{\bf x}}_0 \in \mathcal{X} \subset {\rm I\!R}^p$ a vector of unknown constants and $\boldsymbol{\varepsilon} \in {\rm I\!R}^n$ a vector of (zero mean) residuals. 

The `gmwmx` package allows to estimate functional models for which the $i$-th component of the vector $\mathbf{A} {{\bf x}}_0$ can be described as follows:

\begin{equation}
    \mathbb{E}[\mathbf{Y}_i] = \mathbf{A}_i^T {{\bf x}}_0 = a+b\left(t_{i}-t_{0}\right)+\sum_{h=1}^{2}\left[c_{h} \sin \left(2 \pi f_{h} t_{i}\right)+d_{h} \cos \left(2 \pi f_{h} t_{i}\right)\right] + \sum_{k=1}^{n_{g}} g_{k} H\left(t_{i}-t_{k}\right),
\end{equation}

where $a$ is the initial position at the reference epoch $t_0$, $b$ is the velocity parameter, $c_k$  and $d_k$ are the periodic motion parameters ($h = 1$ and $h = 2$ represent the annual and semi-annual seasonal terms, respectively). The offset terms models earthquakes, equipment changes or human intervention in which $g_k$ is the magnitude of the change at epochs $t_k$, $n_g$ is the total number of offsets, and $H$ is the Heaviside step function. 

### Stochastic model
Regarding the stochastic model, we assume that $\boldsymbol{\varepsilon}_i=\mathbf{Y}_i-\mathbb{E}[\mathbf{Y}_i]$ is a strictly (intrinsically) stationary process and that

\begin{equation}
    \boldsymbol{\varepsilon} \sim \mathcal{F} \left\{\mathbf{0}, \boldsymbol{\Sigma}(\boldsymbol{\gamma}_0)\right\} ,
    \label{eq:model:noise}
\end{equation}

where $\mathcal{F}$ denotes some probability distribution in ${\rm I\!R}^n$ with mean ${\bf 0}$ and covariance $\boldsymbol{\Sigma}(\boldsymbol{\gamma}_0)$. 

We assume that $\boldsymbol{\Sigma}(\boldsymbol{\gamma}_0) > 0$ and that it depends on the unknown parameter vector $\boldsymbol{\gamma}_0 \in \boldsymbol{\Gamma} \subset {\rm I\!R}^q$. This parameter vector specifies the covariance of the observations and is often referred to as the stochastic parameters.

Hence, we let $\boldsymbol{\theta}_0 = \left[\boldsymbol{{\bf x}}_0^{\rm T} \;\; \boldsymbol{\gamma}_0^{\rm T}\right]^{\rm T} \in \boldsymbol{\Theta} = \mathcal{X} \times \boldsymbol{\Gamma} \subset {\rm I\!R}^{p + k}$ denote the unknown parameter vector of the model described above.

The `gmwmx` allows to estimate parameters of a specified functional model as well as parameters of a stochastic model (i.e. $\hat{\boldsymbol{\theta}} = \left[\boldsymbol{\hat{\boldsymbol{x}}}^{T} \;\; \hat{\boldsymbol{\gamma}}^{T}\right]$) defined by a combinations of 

(1) White noise (`wn`)
(2) Matèrn process (`matern`)
(3) Fractional Gaussian noise (`fgn`) and 
(4) Power Law process (`powerlaw`).

## Estimate a model with the GMWMX
You can estimate a model using the GMWMX estimator with the function `estimate_gmwmx()`. 

The stochastic model considered is specified by a string provided to the argument `model_string` which is a combination of the strings `wn`, `powerlaw`, `matern` and `fgn` separated by the character `+`.

You specify the initialization values for solving the optimization problem at the GMWM estimation step that estimate the stochastic model by providing a numeric vector of the correct length (the total number of parameters of the stochastic model specified in `model_string`) to the argument `theta_0`.

You can compute confidence intervals for estimated functional parameters of an estimated model by setting the argument  `ci` to `TRUE`.

Let us consider a single sinusoidal signal with the jumps specified in the `gnssts` object and a combination of a White noise and a Power Law process for the stochastic model. 

```{r}
fit_dobs_wn_plp_gmwmx = estimate_gmwmx(x = data_dobs, theta_0 = c(0.1, 0.1, 0.1), 
                                       model_string = "wn+powerlaw", 
                                       n_seasonal = 1, ci = T)
```

Estimated models are of class `gnsstsmodel`
```{r}
class(fit_dobs_wn_plp_gmwmx)
```


We can print the estimated model or extract estimated parameters (functional and stochastic) as such:
```{r}
print(fit_dobs_wn_plp_gmwmx)
fit_dobs_wn_plp_gmwmx$beta_hat
fit_dobs_wn_plp_gmwmx$theta_hat
```


We can also plot graphically the estimated functional model on the time series and the Wavelet variance of residuals by calling the `plot.gnsstsmodel` method on a `gnsstsmodel` object.
```{r, fig.height=8, fig.width=6, fig.align='center'}
plot(fit_dobs_wn_plp_gmwmx)
```


We can specify the number of iterations of the GMWMX to compute respectively the GMWMX-1 and GMWMX-2 or other iteration of the GMWMX with the argument `k_iter`. For example we can compute the GMWMX-2 as such:

```{r, eval=F, echo=T}
fit_dobs_wn_plp_gmwmx_2 = estimate_gmwmx(x = data_dobs, theta_0 = c(0.1, 0.1, 0.1), model_string = "wn+powerlaw", n_seasonal = 1, k_iter = 2)
```


## Estimate a model with the MLE implemented in Hector

Assuming that you have Hector available on the `PATH`, an estimation of the model can the be performed using the Maximum Likelihood Estimation (MLE) method implemented in [Hector](http://segal.ubi.pt/hector/) as such:

```{r, eval=F, echo=T}
fit_dobs_wn_plp_mle = estimate_hector(x = data_dobs, 
                                      model_string = "wn+powerlaw", 
                                      n_seasonal = 1)




```

```{r, eval=T, echo=F}
file_path_mle = system.file("extdata", "fit_dobs_wn_plp_mle.rda", package = "gmwmx", mustWork = T)
load(file_path_mle)
```

Similarly we can plot and extract the model parameters of the estimated model:
```{r, fig.height=8, fig.width=6, fig.align='center'}
plot(fit_dobs_wn_plp_mle)
fit_dobs_wn_plp_mle$beta_hat
fit_dobs_wn_plp_mle$theta_hat
```

# Importing data from the Plate Boundary Observatory (PBO)

We can load time series data from the [Plate Boundary Observatory (PBO)](https://www.unavco.org/projects/past-projects/pbo/pbo.html) as `gnssts` object  with `PBO_get_station()`:
```{r, eval=F, echo=T}
cola = PBO_get_station("COLA", column = "dE")
```


```{r, eval=F, echo=F}
save(cola, file="cola.rda")
```


```{r, eval=T, echo=F}
cola_path = system.file("extdata", "cola.rda", package = "gmwmx", mustWork = T)
load(cola_path)
```


# Comparing estimated models
Let us consider three potential models for the stochastic model of this signal. More precisely let us consider:

- a White noise and a PowerLaw process 
- a White noise coupled with a Matèrn process
- a White noise and a Fractional Gaussian noise
```{r, fig.height=8, fig.width=6, fig.align='center', eval=T}
fit_cola_wn_plp = estimate_gmwmx(cola, model_string = "wn+powerlaw", 
                                 theta_0 = c(0.1,0.1,0.1),
                                 n_seasonal = 1, 
                                 ci = T)
plot(fit_cola_wn_plp)

fit_cola_wn_fgn = estimate_gmwmx(cola, model_string = "wn+fgn", theta_0 = c(0.1,0.1,0.2),
                                                          n_seasonal = 1, 
                                                          ci = T)
plot(fit_cola_wn_fgn)

fit_cola_wn_matern = estimate_gmwmx(cola, model_string = "wn+matern", 
                                    theta_0 = c(0.1,0.1,0.1,0.1),
                                                          n_seasonal = 1, 
                                                          ci = T)
plot(fit_cola_wn_matern)

```

You can compare estimated models with the function `compare_fits()`
```{r, fig.height=8, fig.width=8, fig.align='center', eval=T}
compare_fits(fit_cola_wn_plp, fit_cola_wn_matern)
```

























