---
title: "Simulation study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval=F,
  comment = "#>"
)
```

This vignette reproduce a design considered in [Cucci et al. (2022)](https://arxiv.org/abs/2206.09668) and implement a Monte Carlo simulation comparing the performance of the GMWMX-1, the GMWMX-2 and the MLE implemented in [Hector](https://teromovigo.com/hector/). Note that this file is not intended to be run on a personal computer but should rather be executed on a high performance computing cluster.

```{r}
rm(list = ls())
```

```{r}
# libraries
library(simts)
library(gmwmx)
library(dplyr)
```


```{r}
check_if_theta_in_ci <- function(theta_vec, ci_mat) {
  vec_emp_coverage <- vector(mode = "logical", length = length(theta_vec))
  for (i in seq(length(theta_vec))) {
    if (dplyr::between(theta_vec[i], ci_mat[i, 1], ci_mat[i, 2])) {
      vec_emp_coverage[i] <- T
    } else {
      vec_emp_coverage[i] <- F
    }
  }
  return(as.numeric(vec_emp_coverage))
}
```


```{r}
# we consider example the model considered in model 3
phase <- 0.45
amplitude <- 2.5
sigma2_wn <- 15
sigma2_powerlaw <- 10
d <- 0.4
bias <- 0
trend <- 5 / 365.25
cosU <- amplitude * cos(phase)
sinU <- amplitude * sin(phase)
```

Let us consider 5 years of daily data
```{r}
# consider n years of daily observations
year <- 20
n <- year * 365
```

We consider a gaussian White noise and a Power Law process as the stochastic model of the residuals $\varepsilon$.
```{r}
# define model for generating gaussian white noise + PLP
model_gaussian_wn_plp <- WN(sigma2 = sigma2_wn) + PLP(sigma2 = sigma2_powerlaw, d = d)
```

We consider three location shifts (jumps) at three different heights.
```{r}
# generate data
# define time at which there are jumps
jump_vec <- c(200, 300, 500)
jump_height <- c(10, 15, 20)
```

We define a seed for reproducibility
```{r}
# define seed
myseed <- 123

# add trend, gaps and sin
nbr_sin <- 1 # define number of sinusoidal process
```

and create the matrix $\boldsymbol{A}$ with function `create_A_matrix()`.
```{r}
# define A matrix
A <- create_A_matrix(1:n, jump_vec, n_seasonal = nbr_sin)
```

We set the vector of functional parameters $\boldsymbol{x}_0$.
```{r}
# define beta
x_0 <- c(bias, trend, cosU, sinU, jump_height)
```

We define the number of simulations to perform
```{r}
# define number of Monte Carlo simulation
n_simu <- 1000

# define number of parameter estimated (depends on the model)
# bias + trend + height * nbr of jump vec + 2* nbr of sin process + wn + powerlaw parameters
nbr_param_check_coverage <- 4
nbr_param <- 2 + length(jump_vec) + 2 * nbr_sin + 3
dim_mat_results <- (nbr_param + nbr_param_check_coverage) * 3 + 1 * 3

# define matrix of results
mat_results <- matrix(NA, ncol = dim_mat_results, nrow = n_simu)
```

We perform the simulation, saving for each Monte Carlo run, the estimated parameters of the functional and stochastic model and the empirical coverage on the functional parameters for the GMWMX-1, the GMWMX-2 as well as the MLE implemented in [Hector](https://teromovigo.com/hector/).
```{r}
for (simu_b in seq(n_simu)) {
  
  # fix seed for reproducibility
  set.seed(myseed + simu_b)

  # generate residuals from a Gaussian White noise + Power law process
  eps <- simts::gen_gts(model = model_gaussian_wn_plp, n = n)

  # create time series
  yy <- A %*% x_0 + eps

  # create gnssts
  gnssts_obj <- create.gnssts(t = 1:length(yy), y = yy, jumps = jump_vec)

  # MLE
  fit_simu_b_mle <- estimate_hector(
    x = gnssts_obj,
    model = "wn+powerlaw",
    n_seasonal = 1
  )
  
  # gmwmx 1 step
  fit_simu_b_gmwm_1_step <- estimate_gmwmx(
    x = gnssts_obj,
    model = "wn+powerlaw",
    theta_0 = c(0.1, 0.1, 0.1),
    n_seasonal = 1,
    ci = T,
    k_iter = 1
  )

  # gmwmx 2 steps
  fit_simu_b_gmwm_2_step <- estimate_gmwmx(
    x = gnssts_obj,
    model = "wn+powerlaw",
    theta_0 = c(0.1, 0.1, 0.1),
    n_seasonal = 1,
    ci = T,
    k_iter = 2
  )


  # define vector of estimators
  fit_mle <- c(fit_simu_b_mle$beta_hat, fit_simu_b_mle$theta_hat)
  fit_gmwm_1_step <- c(fit_simu_b_gmwm_1_step$beta_hat, fit_simu_b_gmwm_1_step$theta_hat)
  fit_gmwm_2_step <- c(fit_simu_b_gmwm_2_step$beta_hat, fit_simu_b_gmwm_2_step$theta_hat)

  # compute coverage for each method
  alpha <- .05
  z_val <- qnorm(1 - alpha / 2)
  
  mat_ci_mle_beta <- matrix(c(
    fit_simu_b_mle$beta_hat - z_val * fit_simu_b_mle$beta_std,
    fit_simu_b_mle$beta_hat + z_val * fit_simu_b_mle$beta_std
  ),
  byrow = F, ncol = 2
  )

  mat_ci_gmwm_1_step_beta <- matrix(c(
    fit_simu_b_gmwm_1_step$beta_hat - z_val * fit_simu_b_gmwm_1_step$beta_std,
    fit_simu_b_gmwm_1_step$beta_hat + z_val * fit_simu_b_gmwm_1_step$beta_std
  ),
  byrow = F, ncol = 2
  )
  
  mat_ci_gmwm_2_step_beta <- matrix(c(
    fit_simu_b_gmwm_2_step$beta_hat - z_val * fit_simu_b_gmwm_2_step$beta_std,
    fit_simu_b_gmwm_2_step$beta_hat + z_val * fit_simu_b_gmwm_2_step$beta_std
  ),
  byrow = F, ncol = 2
  )

  # save empirical coverage
  inside_ci_mle <- check_if_theta_in_ci(x_0, mat_ci_mle_beta)[1:4]
  inside_ci_gmwm_1 <- check_if_theta_in_ci(x_0, mat_ci_gmwm_1_step_beta)[1:4]
  inside_ci_gmwm_2 <- check_if_theta_in_ci(x_0, mat_ci_gmwm_2_step_beta)[1:4]

  # define vector of estimated parameters as object
  res <- c(
    fit_mle, fit_gmwm_1_step,
    fit_gmwm_2_step,
    inside_ci_mle,
    inside_ci_gmwm_1, inside_ci_gmwm_2,
    fit_simu_b_mle$estimation_time[3], fit_simu_b_gmwm_1_step$estimation_time[3], fit_simu_b_gmwm_2_step$estimation_time[3]
  )

  # save in matrix of results
  mat_results[simu_b, ] <- res

  # print status
  cat(paste("Completed simulation", simu_b, "\n",sep = " "))
}
```


```{r}
# define names of mat results
name_param_functionnal <- c("bias", "trend", "cosU", "sinU")
colnames(mat_results) <- c(
    paste("mle", names(fit_mle), sep = "_"),
    paste("gmwm_1_step", names(fit_gmwm_1_step), sep = "_"),
    paste("gmwm_2_step", names(fit_gmwm_2_step), sep = "_"),
    paste0("mle_inside_ci_", name_param_functionnal),
    paste0("gmwm_1_inside_ci_", name_param_functionnal),
    paste0("gmwm_2_inside_ci_", name_param_functionnal),
    c("time_mle", "time_gmwm_1", "time_gmwm_2")
  )
```



